Motional Quantum Ground-State Cooing of a Single Sodium Atom

### 2017-06-10 paper.tex
Update todo
Update results

\begin{abstract}
  Single neutral atom trapped in optical tweezer is a promising system for various applications
  in quantum information and simulation.
  The control of the motional states for such a system is crucial
  for exploiting its full potential.
  We report Raman sideband cooling of a single sodium atom to its three-dimensional
  motional ground state in an optical tweezer.
  Despite having a very large Lamb-Dicke parameter, high initial temperature and
  large differential AC Stark shift in the excited state,
  we achieved a ground state preparation fidelity of $??\%$ after a few hundred ms of cooling
  by using a few new cooling techniques including addressing high order Raman sideband.
  We demonstrate that Raman sideband cooling to the 3D motional ground state is applicable to
  systems where tight confinement or low initial cooling is hard to achieve.
  The is particularly relevant for systems which are challenging to laser-cool,
  such as molecules and exotic atoms, and opens up a new approach for further cooling
  of these systems.
\end{abstract}

@@

Single neutral atoms trapped in optical tweezers provide a promising system
for various applications including quantum information, quantum chemistry
and quantum simulation of many-body systems.
Long-range interactions between the sites can be achieved by using Rydberg atoms,
polar molecules, or dipolar atoms.
Polar molecules offer the additional advantage of a high degree of tunability,
including the shape, range, and strength of the interaction.
Arrays of optical tweezers offer the ability to move, detect, and manipulate individual sites,
and can be used to prepare complex interacting systems with high fidelity.
In order to achieve long coherence time and full quantum control of the system needed for
these applications,
it is usually necessary to control the thermal motion of the atoms by cooling to the
three dimensional motional ground state in the optical tweezer,
which has already been demonstrated with heavy alkali atoms including rubidium and cesium.
However, some of the important properties of previous experiments,
for example low initial temperature and small Lamb-Dicke parameter,
may not be easily achievable for other systems, such as directly laser-cooled polar molecules,
or atoms with challenging sub-Doppler cooling such as Na, K, and Li.
To extend the versatility of this approach, we must be able to apply it in a more general case.
In this letter, we cool single sodium atoms
trapped in optical tweezers to the motional ground state with Raman sideband cooling.
Despite having a large Lamb-Dicke parameter and high initial temperature, we are able to achieve a ground state probability of $XX\%$
by utilizing several new cooling techniques and a carefully optimized cooling sequence.

@@

As mentioned previously, in order to perform Raman sideband cooling efficiently and
minimize the heating during the optical pumping, we need to have a low initial temperature and
a small Lamb-Dicke parameter. However, since the Lamb-Dicke parameter $\eta$ is inversely
proportional to $\sqrt{m}$ (???) where $m$ is the mass of the atom, it is larger for sodium
given the same trap depth (???).
With $45mW$ of power at the focus of the trap, we measured a trapping frequency of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69, 430, 590\}\ \text{\text{kHz}}$
which corresponds to optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.60, 0.24, 0.21\}$.
Moreover, due to the small hyperfine structure in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of $40\mu K$. Combined with the high Lamb-Dicke
parameters, this gives us a initial effective optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_{1eff},\eta^{OP}_{2eff},\eta^{OP}_{3eff}\} = \{2.97, 0.52, 0.40\}$
As a result, there is a very high probability of heating during the optical pumping step
(figure \ref{f-ld}B) causing a $30\%$ average heating probability during optical pumping
in the weakly confined axial direction. Fortunately, the high Lamb-Dicke parameters also
provide us tools to overcome this issue. From our geometry, the Lamb-Dicke parameters for
Raman transitions are $\{\eta^R_{1},\eta^R_{2},\eta^R_{3}\} = \{???, 0.34, 0.29\}$. As shown in
figure \ref{f-ld}A, the high Raman Lamb-Dicke parameters causes a strong coupling to higher orders
of cooling sidebands, especially for high motional states.
This enables us to cool atoms in high motional states by driving on high order Raman sidebands,
removing more motional energy in a single cooling pulse and offsetting the effect of
strong heating. Since the heating probability is higher for high motional states,
cooling on these high order sidebands can greatly suppress the high heating during
optical pumping during the initial cooling. Since the coupling strength of different orders
do not reach their minimums at the same time, using multiple orders of motional sidebands
for cooling also avoids accumulation of population near the coupling minimum of a particular
order, which improves the overall efficiency of the cooling process.

@@

For the more tightly confined radial directions,
we starts the cooling with $\{\bar n_2, \bar n_3\}=2.2, 1.9$.
The radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
where we can clearly see the first order heating, first order cooling and
second order cooling sidebands.
After applying about 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameter is shown in figure \ref{f-radial}B,
where the first and second order cooling sidebands on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $????$ and $????$.
Due to the complexity of the sideband structure,
we measured the Rabi flopping on the carrier and heating sideband before and after cooling
(\ref{f-radial}C and \ref{f-radial}D)
as a independent way to obtain the population in different motional states.
Fitting the Rabi flopping data give a ground state probability of $????$ and $????$,
showing good agreement between the two methods.

@@

Combining the axial and radial cooling results,
we have achieved a ground state probability of $....$.
We believe this is currently limited by off-resonant scattering from the Raman beams
and the resonance shift caused by magnetic field fluctuation.
Improvements on these should improve the cooling performance even further.
We have shown that despite the difficulty in cooling sodium atoms,
it is possible to achieve reliable three dimensional cooling with significant ground state
probability using new cooling techniques and a optimized cooling sequence.
This opens up the possibility of cooling a large variety of systems to their motional ground
state using Raman sideband cooling that is challenging to laser cool
including molecules and exotic atoms.

% Missing:
% ?? Blackman pulse

>>>>

\begin{abstract}
  Single neutral atom trapped in optical tweezer is a promising system for various applications
  in quantum information and simulation.
  The control of the motional states for such a system is crucial
  for exploiting its full potential.
  We report Raman sideband cooling of a single sodium atom to its three-dimensional
  motional ground state in an optical tweezer.
  Despite having a very large Lamb-Dicke parameter, high initial temperature and
  large differential AC Stark shift in the excited state,
  we achieved a ground state preparation fidelity of $77.1(35)\%$ after a few hundred ms of cooling
  by using a few new cooling techniques including addressing high order Raman sideband.
  We demonstrate that Raman sideband cooling to the 3D motional ground state is applicable to
  systems where tight confinement or low initial cooling is hard to achieve.
  The is particularly relevant for systems which are challenging to laser-cool,
  such as molecules and exotic atoms, and opens up a new approach for further cooling
  of these systems.
\end{abstract}

@@

Single neutral atoms trapped in optical tweezers provide a promising system
for various applications including quantum information, quantum chemistry
and quantum simulation of many-body systems.
Long-range interactions between the sites can be achieved by using Rydberg atoms,
polar molecules, or dipolar atoms.
Polar molecules offer the additional advantage of a high degree of tunability,
including the shape, range, and strength of the interaction.
Arrays of optical tweezers offer the ability to move, detect, and manipulate individual sites,
and can be used to prepare complex interacting systems with high fidelity.
In order to achieve long coherence time and full quantum control of the system needed for
these applications,
it is usually necessary to control the thermal motion of the atoms by cooling to the
three dimensional motional ground state in the optical tweezer,
which has already been demonstrated with heavy alkali atoms including rubidium and cesium.
However, some of the important properties of previous experiments,
for example low initial temperature and small Lamb-Dicke parameter,
may not be easily achievable for other systems, such as directly laser-cooled polar molecules,
or atoms with challenging sub-Doppler cooling such as Na, K, and Li.
To extend the versatility of this approach, we must be able to apply it in a more general case.
In this letter, we cool single sodium atoms
trapped in optical tweezers to the motional ground state with Raman sideband cooling.
Despite having a large Lamb-Dicke parameter and high initial temperature, we are able to achieve a ground state probability of $77.1(35)\%$
by utilizing several new cooling techniques and a carefully optimized cooling sequence.

@@

As mentioned previously, in order to perform Raman sideband cooling efficiently and
minimize the heating during the optical pumping, we need to have a low initial temperature and
a small Lamb-Dicke parameter. However, since the Lamb-Dicke parameter $\eta$ is inversely
proportional to $\sqrt{m}$ (???) where $m$ is the mass of the atom, it is larger for sodium
given the same trap depth (???).
With $45mW$ of power at the focus of the trap, we measured a trapping frequency of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69, 430, 590\}\ \text{\text{kHz}}$
which corresponds to optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.60, 0.24, 0.21\}$.
Moreover, due to the small hyperfine structure in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of $40\mu K$. Combined with the high Lamb-Dicke
parameters, this gives us a initial effective optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_{1eff},\eta^{OP}_{2eff},\eta^{OP}_{3eff}\} = \{2.97, 0.52, 0.40\}$
As a result, there is a very high probability of heating during the optical pumping step
(figure \ref{f-ld}B) causing a $30\%$ average heating probability during optical pumping
in the weakly confined axial direction. Fortunately, the high Lamb-Dicke parameters also
provide us tools to overcome this issue. From our geometry, the Lamb-Dicke parameters for
Raman transitions are $\{\eta^R_{1},\eta^R_{2},\eta^R_{3}\} = \{0.40, 0.34, 0.29\}$. As shown in
figure \ref{f-ld}A, the high Raman Lamb-Dicke parameters causes a strong coupling to higher orders
of cooling sidebands, especially for high motional states.
This enables us to cool atoms in high motional states by driving on high order Raman sidebands,
removing more motional energy in a single cooling pulse and offsetting the effect of
strong heating. Since the heating probability is higher for high motional states,
cooling on these high order sidebands can greatly suppress the high heating during
optical pumping during the initial cooling. Since the coupling strength of different orders
do not reach their minimums at the same time, using multiple orders of motional sidebands
for cooling also avoids accumulation of population near the coupling minimum of a particular
order, which improves the overall efficiency of the cooling process.

@@

For the more tightly confined radial directions,
we starts the cooling with $\{\bar n_2, \bar n_3\}=2.2, 1.9$.
The radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
where we can clearly see the first order heating, first order cooling and
second order cooling sidebands.
After applying about 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameter is shown in figure \ref{f-radial}B,
where the first and second order cooling sidebands on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $89.7(19)\%$ and $93.8(26)\%$.
Due to the complexity of the sideband structure,
we measured the Rabi flopping on the carrier and heating sideband before and after cooling
(\ref{f-radial}C and \ref{f-radial}D)
as a independent way to obtain the population in different motional states.
Fitting the Rabi flopping data give a ground state probability of $????$ and $????$,
showing good agreement between the two methods.

@@

Combining the axial and radial cooling results,
we have achieved a ground state probability of $77.1(35)\%$.
We believe this is currently limited by off-resonant scattering from the Raman beams
and the resonance shift caused by magnetic field fluctuation.
Improvements on these should improve the cooling performance even further.
We have shown that despite the difficulty in cooling sodium atoms,
it is possible to achieve reliable three dimensional cooling with significant ground state
probability using new cooling techniques and a optimized cooling sequence.
This opens up the possibility of cooling a large variety of systems to their motional ground
state using Raman sideband cooling that is challenging to laser cool
including molecules and exotic atoms.

% Missing:
% ?? Blackman pulse vs square pulse
